---
title: "Introduction to the BetaDanish Package"
author: "Bilal Ahmad & Dr. Muhammad Yameen Danish"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the BetaDanish Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## 1. Introduction

The **`BetaDanish`** package provides a comprehensive suite of tools for survival and reliability analysis using the Beta-Danish distribution.

Classical lifetime models, such as the Weibull, Gamma, or Log-Normal distributions, are often limited in their hazard shape flexibility. They typically force the hazard rate to be strictly monotonic or, at best, unimodal. The Beta-Danish distribution overcomes these limitations by accommodating **decreasing, increasing, unimodal (upside-down bathtub), and bathtub-shaped** hazard rates within a single, unified four-parameter family.

This package implements the core mathematical functions, robust Maximum Likelihood Estimation (MLE) for complete and right-censored data, and advanced modules for Accelerated Failure Time (AFT) regression, cure-rate models, and competing risks.

## 2. Theoretical Background

The Beta-Danish distribution is constructed by applying the beta-generated transformation (Eugene et al., 2002) to a baseline distribution. The baseline used here is the two-parameter exponentiated log-logistic distribution.

Let $a, b, c, k > 0$ be the parameters of the distribution. 
* $k$ is the **scale parameter**.
* $c$ is the **baseline shape parameter**.
* $a$ and $b$ are the **beta-generator shape parameters** controlling skewness and tail weight.

The Cumulative Distribution Function (CDF) is given by the regularized incomplete beta function:
$$ F(t) = I_{G(t)}(a, b) $$
where $G(t) = \left( \frac{k t}{1 + k t} \right)^c$.

The Probability Density Function (PDF) is:
$$ f(t) = \frac{c k}{B(a, b)} \frac{(k t)^{ca - 1}}{(1 + k t)^{ca + 1}} \left[ 1 - \left( \frac{k t}{1 + k t} \right)^c \right]^{b - 1} $$

### The 3-Parameter Submodel
By fixing $a = 1$, the distribution reduces to a highly tractable 3-parameter submodel (the Complementary Exponentiated Danish distribution), which is particularly useful for regression and cure-rate modeling to ensure parameter identifiability.

## 3. Core Distribution Functions

The package provides the standard `d/p/q/r/h` functions. All internal calculations are performed in log-space to ensure numerical stability, especially in the tails.

```r
library(BetaDanish)

# Calculate the hazard rate at time t = 2
hbetadanish(x = 2, a = 1.5, b = 2.0, c = 3.0, k = 0.5)

# Generate 100 random survival times
set.seed(2026)
sim_data <- rbetadanish(n = 100, a = 1.5, b = 2.0, c = 3.0, k = 0.5)
```

## 4. Fitting Models to Data

The `fit_betadanish()` function is the core MLE engine. It uses a multi-start optimization strategy to ensure global convergence and strictly enforces positivity constraints via log-scale reparameterization.

### Example 1: Uncensored Data (Bladder Cancer Remission)

The `remission` dataset contains the remission times of 128 bladder cancer patients. This data exhibits a unimodal/decreasing hazard rate.

```r
# Load the built-in dataset
data("remission", package = "BetaDanish")

# Fit the full 4-parameter model
fit_full <- fit_betadanish(survival::Surv(time, status) ~ 1, data = remission)
summary(fit_full)

# Fit the 3-parameter submodel (a = 1)
fit_sub <- fit_betadanish(survival::Surv(time, status) ~ 1, data = remission, submodel = TRUE)

# Compare the nested models using a Likelihood Ratio Test (LRT)
compare_models(fit_full, fit_sub)
```

You can easily generate publication-quality diagnostic plots:

```r
# Generates Survival, Hazard, Density, P-P, and Q-Q plots
plot(fit_full, type = "all")
```

### Example 2: Right-Censored Data (Bone Marrow Transplant)

The `transplant` dataset contains survival times for 91 patients, including right-censored observations. The `fit_betadanish()` function natively handles `survival::Surv` objects.

```r
data("transplant", package = "BetaDanish")

# Fit the model to censored data
fit_censored <- fit_betadanish(survival::Surv(time, status) ~ 1, data = transplant)
summary(fit_censored)

# Plot the Kaplan-Meier curve overlaid with the Beta-Danish fit
plot(fit_censored, type = "survival")
```

## 5. Advanced Modeling: Cure Rates

In many clinical datasets (like the `transplant` data), a proportion of patients may be "cured" and will never experience the event. Standard survival models force the survival curve to zero, which misrepresents the data.

The `BetaDanish` package provides the `fit_bd_cure()` function to fit both **Mixture** and **Promotion-Time (Non-Mixture)** cure models.

```r
# Fit a mixture cure model
# Latency (time to event for susceptible) has no covariates (~ 1)
# Incidence (probability of being susceptible) depends on treatment group (~ group)
cure_fit <- fit_bd_cure(
  formula_aft = survival::Surv(time, status) ~ 1,
  formula_cure = ~ group,
  data = transplant,
  type = "mixture"
)

summary(cure_fit)
```

## 6. Automated Benchmarking

To justify the use of the Beta-Danish distribution, you can benchmark it against classical distributions (Weibull, Gamma, Log-Normal, etc.) using the `compare_distributions()` function. *(Note: This requires the `flexsurv` package).* 

```r
# Returns a ranked table of AIC, BIC, and Log-Likelihoods
compare_distributions(fit_full)
```

## 7. Conclusion

The `BetaDanish` package offers a robust, flexible, and user-friendly environment for advanced survival analysis. By unifying standard MLE fitting, comprehensive diagnostics, and advanced modules (AFT, Cure, Competing Risks) under a single framework, it serves as a powerful tool for statisticians and applied researchers alike.

## New in 0.3.0

```{r bd_intro_setup, include = FALSE}
# The blocks above this point are ```r display blocks, which knitr does
# not execute, so the package has not been attached yet.
library(BetaDanish)
```

Three groups of functions were added after the 0.2.0 release.

### Structural properties

```{r}
p <- list(a = 1.5, b = 5, c = 2, k = 1)

bd_moment_summary(p$a, p$b, p$c, p$k)
bd_entropy_shannon(p$a, p$b, p$c, p$k)
bd_stress_strength(strength = p, stress = p)   # identical laws: one half
```

Moments exist only when `b > r`, and the package enforces it rather than
returning a plausible-looking number from a truncated sum:

```{r}
bd_moments(1:4, a = 1.5, b = 3, c = 2, k = 1)   # the fourth is Inf
bd_tail_index(a = 1.5, b = 3, c = 2, k = 1)$moment_condition
```

### Shape, before and after fitting

`bd_ttt_plot()` reads the hazard shape from the data alone, with no model
assumed. `bd_hazard_shape()` does the same from a fitted parameter set, so
the two can be compared.

```{r, fig.width = 6, fig.height = 4}
data(guinea_pig)
ttt <- bd_ttt_plot(guinea_pig$time)
attr(ttt, "shape")
```

### Interval estimation

All four parameters are positive, so a symmetric Wald interval on the
natural scale can cross zero. `bd_wald_ci()` builds on the log scale
instead, and `bd_profile_ci()` inverts the likelihood ratio test. When the
profile never returns below the critical threshold there is no finite upper
bound, and the honest report is a lower bound rather than a point estimate
with an interval around it.

```{r, eval = FALSE}
fit <- fit_betadanish(survival::Surv(time, status) ~ 1, data = remission,
                      submodel = TRUE)
bd_wald_ci(fit)
bd_profile_plot(bd_profile_ci(fit, "b"))

# For the four-parameter model, report the identified composite
bd_identified_coef(fit_betadanish(survival::Surv(time, status) ~ 1,
                                  data = remission))
```

### Simulation studies

`bd_simulation_study()`, `bd_simulation_cure()` and
`bd_simulation_competing()` report bias, RMSE, the ratio of mean estimated
standard error to empirical standard deviation, and Wald coverage. Read the
last two together: a ratio well below one means the asymptotic standard
errors are optimistic at that sample size, and coverage falls short as a
consequence.

```{r, eval = FALSE}
bd_simulation_study(n = c(50, 100, 200), n_sim = 500,
                    truth = c(b = 3, c = 2, k = 0.5), submodel = TRUE)
```
